---
title: "Laboratory with R"
author: "Enrique Akira Tagikuchi e Itziar Salazar"
date: "2025-11-29"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

### Download data (1a, 1b)

```{r Download}
source('CreateDF.R')
if (!file.exists('sea_data.RData')){
  download_data('sea_data.RData')
}
load('sea_data.RData')

```

### Check dimensionality (1c)

```{r, Dimensionality}
cat("Dimension df_yearly\n")
dim_sea.deep <- dim(sea.deep)
cat("Rows (obs.):", dim_sea.deep[1], "\n")
cat("Columns (var.):", dim_sea.deep[2], "\n")
print(str(sea.deep)) #Structure of the df, dimensions, variables and data types
print(summary(sea.deep)) #Descriptive statistical summary: mean, quantiles, min, max, NAs...
print(sum(is.na(sea.deep))) #Just to make sure 
```

### Boxplot representation (1d)

```{r Boxplots, echo=FALSE}
library(ggplot2)
ggplot(sea.deep, aes(x = factor(fondària), y = temperatura)) +
  geom_boxplot() +
  facet_wrap(~any) +
  labs(
    title = "Boxplot de la temperatura per fondària (2000-2017)",
    x = "Fondària (m)", 
    y = "Temperatura (°C)"
  )
```

### Statistical analysis (1e)

```{r, StatisticalAnalysis}
library(dplyr)

stats1 <- sea.deep %>%
  group_by(fondària, any) %>%
  summarise(
    Mitjana = mean(temperatura),
    Mediana = median(temperatura),
    SD = sd(temperatura),
    IQR = IQR(temperatura),
    Min = min(temperatura), #Other statistics of interest
    Max = max(temperatura),
    Range = Max - Min, #Difference between max and min values
    CV = SD / Mitjana, #Coefficient of Variation (CV)
    .groups = 'drop'
   ) 

print(stats1) 
View(stats1) #Average temperature per depth per year

temp_dy <- sea.deep %>% 
  group_by(fondària, any) %>% #Variables of interest
  summarise(
    temp_mitja = mean(temperatura),
    .groups = 'drop'
  ) #New variable average temperature per depth per year
print(temp_dy)

stats2 <- temp_dy %>%
  group_by(fondària) %>%
  summarise(
    Mitjana = mean(temp_mitja),
    Mediana = median(temp_mitja),
    SD = sd(temp_mitja),
    IQR = IQR(temp_mitja),
    Min = min(temp_mitja), #Other statistics of interest
    Max = max(temp_mitja),
    Range = Max - Min, #Difference between max and min values
    CV = SD / Mitjana, #Coefficient of Variation (CV)
    .groups = 'drop'
  )

print(stats2) 
View(stats2) #Average temperature per depth across 2000-2017   

```

### Annual variations (1f)

```{r AnnnualVariations, echo=FALSE}
stats1 <- stats1 %>%
  mutate(fondària = factor(fondària, levels = c(0, -20, -50, -80)))

ggplot(stats1, aes(
  x = any, 
  y = Mitjana, 
  color = fondària)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Variació anual de la temperatura mitjana per fondària",
    subtitle = " (2000-2017) Mar Mediterrani, punt d'observació: 42º 03' N, 3º 15' E", #####Acabar de decidir títols i subtítols
    x = "Any",
    y = "Temperatura mitjana (°C)",
    color = "Fondària (m)"
  ) +
  theme_minimal()

```

### Save to Excel (1g)

```{r, Excel}
library(openxlsx)

new_var <- createWorkbook()
#Average temperature per depth per year
addWorksheet(new_var, "Stats_per_any")
writeData(new_var, "Stats_per_any", stats1)
#Average temperature per depth across 2000-2017   
addWorksheet(new_var, "Stats_globals")
writeData(new_var, "Stats_globals", stats2)
#Save on a new excel file
saveWorkbook(new_var, "NUEVO.xlsx", overwrite = TRUE)
```

## Exercise 2 

### Plot1 - Heatmap

```{r Heatmap, echo=FALSE}
#Executo els paquets a utilitzar 
library(tidyr)
library(gplots)
library(RColorBrewer)

#Elimino observacions "mitjana anual", conservo només els mesos Gener-Desembre
if ("mes" %in% names(sea.deep)) {
  obs <- sea.deep[sea.deep$mes != "Mitjana anual", ]
} else {
  obs <- sea.deep
}

temp_dy <- obs %>% #Calculo la temperatura mitjana per any per fondària 
  group_by(any, fondària) %>%
  summarise(
    mean = mean(temperatura),
    sd   = sd(temperatura),
    n    = n()
  ) %>%
  ungroup() %>%
  arrange(any, fondària)

#Preparo les dades (el·laboració de matius) per a la representació heatmap.2 
fondària <- c(-80, -50, -20, 0)

matriu_1 <- temp_dy %>%
  select(any, fondària, mean) %>%
  pivot_wider(names_from = fondària, values_from = mean) %>%
  arrange(any)
anys <- matriu_1$any
columnes <- intersect(as.character(fondària), colnames(matriu_1))
rownames(matriu_1) <- anys
mitjana_temp_mat <- as.matrix(matriu_1[, columnes])

matriu_2 <- temp_dy %>%
  select(any, fondària, sd) %>%
  pivot_wider(names_from = fondària, values_from = sd) %>%
  arrange(any)
rownames(matriu_2) <- matriu_2$any
sd_temp_mat <- as.matrix(matriu_2[, columnes])

tons_heatmap <- colorRampPalette(brewer.pal(9, "YlOrRd"))(100) #Selecciono tonalitat mapa

heatmap.2( #Fem el mapa
  mitjana_temp_mat,
  Rowv = FALSE, Colv = FALSE, dendrogram = "none",
  trace = "none", key = TRUE, density.info = "none",
  col = tons_heatmap, na.color = "grey90",
  margins = c(7, 9),
  main = "",
  xlab = "Fondària (m)", ylab = "Any",
  labCol = columnes,
  labRow = as.character(anys),
  cexRow = 0.9, cexCol = 1.0,
  add.expr = {
    nr <- nrow(sd_temp_mat); nc <- ncol(sd_temp_mat)
    for (j in seq_len(nc)) {
      s <- sd_temp_mat[, j]
      if (all(is.na(s))) next
      rng <- range(s)
      y <- sapply(seq_len(nr), function(i) nr - i + 1)
      if (!is.finite(diff(rng)) || diff(rng) == 0) {
        x <- rep(j, nr)
      } else {
        x_scaled <- (s - rng[1]) / diff(rng)
        x <- (j - 0.5) + x_scaled
      }
      lines(x, y, col = "cyan3", lwd = 2)
      points(x, y, col = "cyan3", pch = 20, cex = 0.75)
    }
  }
)
#Incorporo el títol més tard i l'adjusto per evitar encabalcaments
title("Temperatura mitjana per any i fondària",
      adj = 1,           
      line = 1.2,        
      cex.main = 1.1)   
```

######The heatmap sh

ows how the annual mean water temperature changes at different depths (-80, -50, -20 and 0 meters) between 2000 and 2017. The colour distribution shows that surface waters (0 and -20 meters) are consistently the warmest and exhibit the greatest interannual variability. This variability is evident in the blue lines representing the standard deviation (SD), which are imposed on the heatmap. The deeper layers (-50 and -80 meters) display a colder and more stable thermal profile, with significantly reduced standard deviation, indicating lower influence from seasonal cycles. The period between 2014 and 2016 stands out as particularly intense warming conditions are observed at the surface, potentially driven by large-scale oceanographic or climatic dynamics.

### Plot2 - 

```{r Radial}
library(plotrix)

mes_l <- c("Gener", "Febrer", "Març", "Abril", "Maig", "Juny", 
           "Juliol", "Agost", "Setembre", "Octubre", "Novembre", "Desembre")

radial_data <- sea.deep %>%
  filter(mes != 'Mitjana anual') %>%
  mutate(mes = factor(mes, levels = mes_l)) %>%
  group_by(mes, fondària) %>%
  summarise(avg_temp = mean(temperatura, na.rm = TRUE), .groups = 'drop')

plot_matrix <- radial_data %>%
  pivot_wider(names_from = fondària, values_from = avg_temp) %>%
  arrange(mes) %>%
  select(-mes) %>%
  as.matrix()

plot_matrix = rbind(plot_matrix, plot_matrix[1, 1:4])

depth_levels <- colnames(plot_matrix)
num_depths <- length(depth_levels)

col_set = c("blue","cyan", "green", "coral")

angles <- c((0:(length(mes_l) - 1)) * (2 * pi / length(mes_l)),0)

radial.plot(
  t(plot_matrix), radial.pos = t(replicate(4, angles)), rp.type = "l",
  lty = 4, lwd = 2,
  labels = mes_l, label.pos = angles, line.col = col_set,
  radial.lim = c(0, 30),
  main = "Average Monthly Temperature by Depth (All Years)",
  clockwise = TRUE, start = pi/2
)


legend(
  "bottomright", legend = paste(depth_levels, "m"),
   col=col_set, lty = 4, lwd = 2, bty = "n"
)
```

## Exercise 3

```{r MontlyChange}
library(stringr )
source('Ex3.R')
plots = ex3(sea.deep, sea.pred)
print(plots$a)

```

```{r Historical}
print(plots$b1)
```

```{r YearlyHistorical}
print(plots$b2)
```
